Computing Entropies With Nested Sampling

### 2017-05-21 paper/ms.tex
Bit of writing

If an unknown quantity $x$ has a discrete probability distribution $p(x)$,
the Shannon entropy \citep{shannon} is defined as
\begin{align}
H_{x} &= -\sum_{x} p(x) \log p(x)
\end{align}
where the sum is over all of the possible values of $x$ under consideration.

The entropy quantifies the degree to which the issue
``what is the value of $x$, precisely?'' remains unresolved. For example,
if there is only one possible value which has probability 1, $H_x$ is
zero\footnote{Conventionally, $0 \times \log 0$ is defined to be equal
to $\lim_{x \to 0^+} (x\log x)$, which is zero.}.
If there are $N$ possibilities with equal probabilities $1/N$ each,
then the entropy is $H_x = \log(N)$.
If the space of possible $x$-values is continuous, the differential
entropy
\begin{align}
H_{x} &= -\int p(x) \log p(x) \, dx
\end{align}
quantifies uncertainty by generalising the log-volume
of the plausible region, where volume is defined with respect to $dx$.

@@

Bayesian inference is the use of probability theory to
describe uncertainty, often about unknown quantities
(``parameters'') $\x$. Some data $\d$, initially unknown but
thought to be relevant to $\x$, is obtained.
The prior information leads the
user to specify a prior distribution $p(\x)$ for the unknown parameters,
along with a ``sampling distribution'' $p(\d | \x)$ describing prior knowledge
of how the data is related to the parameters
(i.e., if the parameters where known, what data would likely be observed?).
By the product rule, this yields a {\em joint prior}
\begin{align}
p(\x, \d) &= p(\x)p(\d | \x)
\end{align}
which is the starting point for
Bayesian inference \citep{caticha2008lectures}.

In practical applications, the following operations
are usually feasible and have a low computational cost:
\begin{enumerate}
  \item Points can be generated from the prior $p(\x)$;
  \item Simulated datasets can be generated from $p(\d | \x)$ for any
        given value of $\x$;
  \item The likelihood, $p(\d | \x)$, can be evaluated cheaply for any
        $\d$ and $\x$.
\end{enumerate}
Throughout this paper I assume these operations are available and inexpensive.

@@

The entropy of the prior $p(\x)$ describes the degree to which the question
``what is the value of $\x$, precisely?'' remains unanswered, while the
entropy of the joint prior $p(\x, \d)$
describes the degree to which the question
``what is the value of the pair $(\x, \d)$, precisely?'' remains unanswered.

The degree to which the question ``what is the value of $\x$?'' is
answered by the question ``what is the value of $\d$?'' is given by the
conditional entropy
\begin{align}
H_{\x | \d} &= - \sum_{\d} p(\d) \sum_{\x} p(\x | \d) \log p(\x | \d)
\end{align}
which is the expected value of the entropy of the posterior, averaged over
all possible datasets which might be observed.

Suppose we wanted to compute $H_{\x | \d}$.
This would be difficult because the expression
for the posterior distribution
\begin{align}
p(\x | \d) &= \frac{p(\x)p(\d | \x)}{p(\d)}
\end{align}
contains the marginal likelihood integral:
\begin{align}
p(\d) &= \int p(\x) p(\d | \x) \, d\x
\end{align}
also known as the `evidence'. In this paper I use the term evidence because
for brevity.

It is important to distinguish between $H_{\x | \d}$ and the
entropy of $\x$ given a particular value of $\d$, which might be written
$H_{\x | \d=\d_0}$.
The former measures the degree to which one question answers
another (ex ante), so is a function of two questions. The latter measures the
degree to which a {\em statement} answered a question (ex post),
and is a function of a question and a statement.

@@

For experimental design purposes, maximising either $I_{\x, \d}$ or
$H_{\x | \d}$ will produce the same result. {\todo Mention reference priors?}

@@

@@

Continuous hypothesis spaces are commonly used in practice. Suppose
$\x$ has probability density $f(\x)$. The algorithm actually estimates
\begin{align}
h &= \int f(\x) \int f(\x')
            \log \left[ P(d(\x', \x) < L | \x) \right]
                        \, d\x' \, d\x.
\end{align}
If we actually want the differential
entropy
\begin{align}
h_{\x} &= -\int f(x) \log f(x) \, dx
\end{align}
we can use the fact that density equals mass divided
by volume. Assume $L$ is small, so that
\begin{align}
P(d(\x', \x) < L | \x)
    &\approx
    f(\x) \int_{d(\x', \x) < L} \, d\x'\\
    &= f(\x) \, V
\end{align}
where $V$ is the volume of the region where $d(\x', \x) < L$.

>>>>

If an unknown quantity $x$ has a discrete probability distribution $p(x)$,
the Shannon entropy \citep{shannon} is defined as
\begin{align}
H_{x} &= -\sum_{x} p(x) \log p(x)
\end{align}
where the sum is over all of the possible values of $x$ under consideration.
The entropy quantifies the degree to which the issue
``what is the value of $x$, precisely?'' remains unresolved
\citep{knuth_questions}. For example,
if there is only one possible value which has probability 1, $H_x$ is
zero\footnote{Conventionally, $0 \times \log 0$ is defined to be equal
to $\lim_{x \to 0^+} (x\log x)$, which is zero.}.
If there are $N$ possibilities with equal probabilities $1/N$ each,
then the entropy is $H_x = \log(N)$.
If the space of possible $x$-values is continuous, the differential
entropy
\begin{align}
H_{x} &= -\int p(x) \log p(x) \, dx
\end{align}
quantifies uncertainty by generalising the log-volume
of the plausible region, where volume is defined with respect to $dx$.

@@

Bayesian inference is the use of probability theory to
describe uncertainty, often about unknown quantities
(`parameters') $\x$. Some data $\d$, initially unknown but
thought to be relevant to $\x$, is obtained.
The prior information leads the
user to specify a prior distribution $p(\x)$ for the unknown parameters,
along with a ``sampling distribution'' $p(\d | \x)$ describing prior knowledge
of how the data is related to the parameters
(i.e., if the parameters where known, what data would likely be observed?).
By the product rule, this yields a {\em joint prior}
\begin{align}
p(\x, \d) &= p(\x)p(\d | \x)
\end{align}
which is the starting point for
Bayesian inference \citep{caticha2008lectures}.
In practical applications, the following operations
are usually feasible and have a low computational cost:
\begin{enumerate}
  \item Points can be generated from the prior $p(\x)$;
  \item Simulated datasets can be generated from $p(\d | \x)$ for any
        given value of $\x$;
  \item The likelihood, $p(\d | \x)$, can be evaluated cheaply for any
        $\d$ and $\x$.
\end{enumerate}
Throughout this paper I assume these operations are available and inexpensive.

@@

The entropy of the prior $p(\x)$ describes the degree to which the question
``what is the value of $\x$, precisely?'' remains unanswered, while the
entropy of the joint prior $p(\x, \d)$
describes the degree to which the question
``what is the value of the pair $(\x, \d)$, precisely?'' remains unanswered.
The degree to which the question ``what is the value of $\x$?'' would remain
unresolved if $\d$ were resolved is given by the
conditional entropy
\begin{align}
H_{\x | \d} &= - \sum_{\d} p(\d) \sum_{\x} p(\x | \d) \log p(\x | \d)
\end{align}
which is the expected value of the entropy of the posterior, averaged over
all possible datasets which might be observed.

Suppose we wanted to compute $H_{\x | \d}$, perhaps to compare it to
$H_{\x}$ and quantify how much might be learned about $\x$.
This would be difficult because the expression
for the posterior distribution
\begin{align}
p(\x | \d) &= \frac{p(\x)p(\d | \x)}{p(\d)}
\end{align}
contains the marginal likelihood integral:
\begin{align}
p(\d) &= \int p(\x) p(\d | \x) \, d\x
\end{align}
also known as the `evidence', which tends to be computable but costly.

It is important to distinguish between $H_{\x | \d}$ and the
entropy of $\x$ given a particular value of $\d$, which might be written
$H_{\x | \d=\d_0}$ (in traditional notation, $H(X|Y)$ and $H(X|Y=y)$).
The former measures the degree to which one question answers
another (ex ante), so is a function of two questions. The latter measures the
degree to which a {\em statement} answered a question (ex post),
and is a function of a question and a statement.

@@

For experimental design purposes, if we want to maximise the expected amount
of information obtained from data, maximising either $I_{\x, \d}$ or
$H_{\x | \d}$ will produce the same result because the prior $p(\x)$ does not
vary with the experimental design. Reference priors {\bf cite Bernardo}
also make use of the mutual information but in a different way.

@@

We would like to calculate the expected value of a log-probability:

@@

Continuous hypothesis spaces are commonly used in practice. Suppose
$\x$ has probability density $f(\x)$, so the differential entropy is
\begin{align}
H &= -\int_{-\infty}^{\infty} f(\x) \log f(\x) \, d\x
\end{align}

What the algorithm actually estimates is
\begin{align}
h &= \int f(\x) \int f(\x')
            \log \left[ P(d(\x', \x) < L | \x) \right]
                        \, d\x' \, d\x.
\end{align}
If we actually want the differential
entropy, we can use the fact that density equals mass divided
by volume. Assume $L$ is small, so that
\begin{align}
P(d(\x', \x) < L | \x)
    &\approx
    f(\x) \int_{d(\x', \x) < L} \, d\x'\\
    &= f(\x) \, V
\end{align}
where $V$ is the volume of the region where $d(\x', \x) < L$.
